Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
finish filling in content from planning notes

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'ESSD')`"
author:
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "2" 
  - given_name: Ben 
    surname: Bond-Lamberty (confirm)
    affiliation: "3" 
  - given_name: Susan 
    surname: Cook-Patton (confirm)
    affiliation: "4" 
  - given_name: Sandro 
    surname: Federici (confirm)
    affiliation: "5" 
  - given_name: Helene
    surname: Muller-Landau
    affiliation: "6" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "7" 
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Valentyna
    surname: Slivinska (confirm)
    affiliation: "5"
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,6 *"
    email: teixeirak@si.edu
    corresponding: true

@@

affiliation:
  - code: 1
    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
  - code: 2
    address: 
  - code: 3
    address: 
  - code: 4
    address: 
  - code: 5
    address: 
  - code: 6
    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
  - code: 7
    address: CIRAD, Montpellier, France

abstract: |
  The abstract goes here.
  It can also be on _multiple lines_.
bibliography: references.bib
running:
  title: IPCC forest C accounting with ForC
  author: Williams et al.
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company. 
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #code: |
  #  use this to add a statement when having only software code available
  #data: |
  #  use this to add a statement when having only data sets available
  codedata: |
    use this to add a statement when having data sets and software code available
  sample: |
    use this section when having geoscientific samples available
videosupplement: |
  use this section when having video supplements available
authorcontribution: |
  Daniel wrote the package. Josiah thought about poterry. Markus filled in for a second author.
disclaimer: |
  We like Copernicus.
acknowledgements: |
  Thanks to the rticles contributors!
appendix: |
  \section{Figures and tables in appendices}
  Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
  \subsection{Option 1}
  If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
  They will be correctly named automatically.
  \subsection{Option 2}
  If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.

  To rename them correctly to A1, A2, etc., please add the following commands in front of them:
  `\appendixfigures` needs to be added in front of appendix figures
  `\appendixtables` needs to be added in front of appendix tables

  Please add `\clearpage` between each table and/or figure. Further guidelines on figures and tables can be found below.
output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

The following settings can or must be configured in the header of this file and are bespoke for Copernicus manuscripts:

- The `journal` you are submitting to using the official abbreviation. You can use the function `rticles::copernicus_journal_abbreviations(name = '...')` to search the existing journals.

- Specific sections of the manuscript:

  - `running` with `title` and `author`

  - `competinginterests`

  - `copyrightstatement` (optional)

  - `availability` (strongly recommended if any used), one of `code`, `data`, or `codedata`

  - `authorcontribution`

  - `disclaimer`

  - `acknowledgements`

See the defaults and examples from the skeleton and the official Copernicus documentation for details.

@@

*(Importance of forests for climate change mitigation)*

*(Need for good data in international carbon accounting)*

*(Introduce EFDb & forc)*

Example citation [@anderson-teixeira_forc_2018]

Here, we:
(1) clarify C cycle terminology
(2) describe mapping of ForC to IPCC's EFDB,
(3) describe updates to ForC  (ForC v4.0)
(4) summarize the data in ForC that's relevant to EFDB, identifying gaps
(5) provide recommendations for improving data collection, analysis, database, and accounting

@@

>>>>

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'ESSD')`"
author:
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "2" 
  - given_name: Ben 
    surname: Bond-Lamberty (confirm)
    affiliation: "3" 
  - given_name: Susan 
    surname: Cook-Patton
    affiliation: "4" 
  - given_name: Sandro 
    surname: Federici 
    affiliation: "5" 
  - given_name: Helene
    surname: Muller-Landau
    affiliation: "6" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "7" 
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Valentyna
    surname: Slivinska 
    affiliation: "5"
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,6 *"
    email: teixeirak@si.edu
    corresponding: true

@@

affiliation:
  - code: 1
    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
  - code: 2
    address: 
  - code: 3
    address: 
  - code: 4
    address: 
  - code: 5
    address:  IPCC Task Force on National Greenhouse Gas Inventories Technical Support Unit, Institute for Global Environmental Strategies, Hayama, Japan
  - code: 6
    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
  - code: 7
    address: CIRAD, Montpellier, France

abstract: |
  The abstract goes here.
  It can also be on _multiple lines_.
bibliography: references.bib
running:
  title: IPCC forest C accounting with ForC
  author: Williams et al.
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company. 
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #code: |
  #  use this to add a statement when having only software code available
  #data: |
  #  use this to add a statement when having only data sets available
  codedata: |
    use this to add a statement when having data sets and software code available
authorcontribution: |
  (fill this in)
disclaimer: |
acknowledgements: |
  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
  Funding for this study was provided by The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
appendix: |

output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

@@

**(Forest are critical for climate change mitigation)**

**(Need for good data in international carbon accounting)**

**(Introduce EFDB & ForC)**

Here, we:
(1) clarify definitions of relevant carbon stocks and increments
(2) describe mapping of ForC to IPCC's EFDB,
(3) describe updates to ForC  (ForC v4.0),
(4) summarize the data in ForC that's relevant to EFDB, identifying gaps, and
(5) provide recommendations for improving data collection, analysis, database, and accounting.
We note that

@@

# Mapping ForC to EFDB

## Carbon cycle variables

**Table: variable  mapping and equations-- give equations to calculate IPCC variables from C cycle variables**

Define relationship among NEE, NEP, and delta.C., especially noting role of harvest.

## Land use categories

# Updates to ForC  (ForC v4.0)

To support export of data to EFDB, and to improve the overall quality of the ForC database, we implemented some modest restructuring, resolved duplicate records, and conducted quality control.

This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].

## ForC restructuring

**Table: changes to ForC fields**.  
Use  this table : https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv.

## Resolving duplicates

## Quality control

# Results

**figure: map of relevant ForC data with underlying FAO ecozones**

summarize the data in ForC that's relevant to EFDB, identifying gaps

dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)

# Recommendations

## Data collection and analysis needs

**(Paragraph highlighting important gaps in variables / regions)**

Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data. 
We recommend that researchers calculate and report these, as specified below.
Furthermore, there is an opportunity to fill data gaps by calculating these from existing census data.
For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; REFS] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available (n=1 for standing dead wood, n=2 for woody mortality and delta.agb) [but see REFS].

**A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.**
*(Camille/Helene can write this paragraph easily.)*

## Data reporting needs

**We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards:**

  - adopt common standards for variables like min diameter of deadwood, select soil sampling increments to include a cutoff at 30.

  - report 95% CIs, SE, or STD and n

  - report C variables in article text, table, or SI table. EFDB cannot accept data digitized from figures

## Database needs

There are plenty of relevant, published data that are not included in ForC. 
Systematic review of the literature could vastly improve data coverage.
*(There are some efforts underway, including a few that Susan can specify.)*

## IPCC

An important challenge is that forests are changing rapidly, and data collected a decaade ago may no longer be relevant, particularly in the cases of C increments and fluxes.

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
finish filling in content from planning notes

\Author[1]{Madison}{Williams}
\Author[1]{Valentine}{Herrmann}
\Author[2]{Rebecca}{Banbury Morgan}
\Author[3]{Ben}{Bond-Lamberty (confirm)}
\Author[4]{Susan}{Cook-Patton (confirm)}
\Author[5]{Sandro}{Federici (confirm)}
\Author[6]{Helene}{Muller-Landau}
\Author[7]{Camille}{Piponiot}
\Author[1]{Teagan}{Rogers}
\Author[5]{Valentyna}{Slivinska (confirm)}
\Author[1,6 *]{Kristina J.}{Anderson-Teixeira}

\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
Biology Institute, Front Royal, VA, United States}
\affil[2]{}
\affil[3]{}
\affil[4]{}
\affil[5]{}
\affil[6]{Forest Global Earth Observatory, Smithsonian Tropical Research
Institute, Panama, Republic of Panama}
\affil[7]{CIRAD, Montpellier, France}

@@

The following settings can or must be configured in the header of this
file and are bespoke for Copernicus manuscripts:

\begin{itemize}
\item
  The \texttt{journal} you are submitting to using the official
  abbreviation. You can use the function
  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
  to search the existing journals.
\item
  Specific sections of the manuscript:

  \begin{itemize}
  \item
    \texttt{running} with \texttt{title} and \texttt{author}
  \item
    \texttt{competinginterests}
  \item
    \texttt{copyrightstatement} (optional)
  \item
    \texttt{availability} (strongly recommended if any used), one of
    \texttt{code}, \texttt{data}, or \texttt{codedata}
  \item
    \texttt{authorcontribution}
  \item
    \texttt{disclaimer}
  \item
    \texttt{acknowledgements}
  \end{itemize}
\end{itemize}

See the defaults and examples from the skeleton and the official
Copernicus documentation for details.

@@

\emph{(Importance of forests for climate change mitigation)}

\emph{(Need for good data in international carbon accounting)}

\emph{(Introduce EFDb \& forc)}

Example citation \citep{anderson-teixeira_forc_2018}

Here, we: (1) clarify C cycle terminology (2) describe mapping of ForC
to IPCC's EFDB, (3) describe updates to ForC (ForC v4.0) (4) summarize
the data in ForC that's relevant to EFDB, identifying gaps (5) provide
recommendations for improving data collection, analysis, database, and
accounting

@@

@@

\sampleavailability{use this section when having geoscientific samples
available} %% use this section when having geoscientific samples available

\videosupplement{use this section when having video supplements
available} %% use this section when having geoscientific samples available

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Figures and tables in appendices}

Regarding figures and tables in appendices, the following two options
are possible depending on your general handling of figures and tables in
the manuscript environment:

\subsection{Option 1}

If you sorted all figures and tables into the sections of the text,
please also sort the appendix figures and appendix tables into the
respective appendix sections. They will be correctly named
automatically.

\subsection{Option 2}

If you put all figures after the reference list, please insert appendix
tables and figures after the normal tables and figures.

To rename them correctly to A1, A2, etc., please add the following
commands in front of them: \texttt{\textbackslash{}appendixfigures}
needs to be added in front of appendix figures
\texttt{\textbackslash{}appendixtables} needs to be added in front of
appendix tables

Please add \texttt{\textbackslash{}clearpage} between each table and/or
figure. Further guidelines on figures and tables can be found below.
\noappendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{Daniel wrote the package. Josiah thought about
poterry. Markus filled in for a second author.} %% optional section

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\disclaimer{We like Copernicus.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
Thanks to the rticles contributors!
\end{acknowledgements}

>>>>

\Author[1]{Madison}{Williams}
\Author[1]{Valentine}{Herrmann}
\Author[2]{Rebecca}{Banbury Morgan}
\Author[3]{Ben}{Bond-Lamberty (confirm)}
\Author[4]{Susan}{Cook-Patton}
\Author[5]{Sandro}{Federici}
\Author[6]{Helene}{Muller-Landau}
\Author[7]{Camille}{Piponiot}
\Author[1]{Teagan}{Rogers}
\Author[5]{Valentyna}{Slivinska}
\Author[1,6 *]{Kristina J.}{Anderson-Teixeira}

\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
Biology Institute, Front Royal, VA, United States}
\affil[2]{}
\affil[3]{}
\affil[4]{}
\affil[5]{IPCC Task Force on National Greenhouse Gas Inventories
Technical Support Unit, Institute for Global Environmental Strategies,
Hayama, Japan}
\affil[6]{Forest Global Earth Observatory, Smithsonian Tropical Research
Institute, Panama, Republic of Panama}
\affil[7]{CIRAD, Montpellier, France}

@@

@@

\textbf{(Forest are critical for climate change mitigation)}

\textbf{(Need for good data in international carbon accounting)}

\textbf{(Introduce EFDB \& ForC)}

Here, we: (1) clarify definitions of relevant carbon stocks and
increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
updates to ForC (ForC v4.0), (4) summarize the data in ForC that's
relevant to EFDB, identifying gaps, and (5) provide recommendations for
improving data collection, analysis, database, and accounting. We note
that

@@

\section{Mapping ForC to EFDB}

\subsection{Carbon cycle variables}

\textbf{Table: variable mapping and equations-- give equations to
calculate IPCC variables from C cycle variables}

Define relationship among NEE, NEP, and delta.C., especially noting role
of harvest.

\subsection{Land use categories}

\section{Updates to ForC (ForC v4.0)}

To support export of data to EFDB, and to improve the overall quality of
the ForC database, we implemented some modest restructuring, resolved
duplicate records, and conducted quality control.

This section describes changes relative to ForC v2.0
\citep{anderson-teixeira_forc_2018}.

\subsection{ForC restructuring}

\textbf{Table: changes to ForC fields}.\\
Use this table :
https://github.com/forc-db/ForC/blob/master/database\_management\_records/record\_of\_changes.csv.

\subsection{Resolving duplicates}

\subsection{Quality control}

\section{Results}

\textbf{figure: map of relevant ForC data with underlying FAO ecozones}

summarize the data in ForC that's relevant to EFDB, identifying gaps

dead wood and litter comparisons will be particularly interesting, as
IPCC values are based on just a handful of references for each climate
zone (table 2.2 in 2019 guidelines)

\section{Recommendations}

\subsection{Data collection and analysis needs}

\textbf{(Paragraph highlighting important gaps in variables / regions)}

Several variables of value to IPCC, including standing dead wood, woody
mortality, delta.agb, are not calculated and presented as frequently as
are AGB and ANPP\_woody, even though they can readily be derived from
the same census data. We recommend that researchers calculate and report
these, as specified below. Furthermore, there is an opportunity to fill
data gaps by calculating these from existing census data. For example,
the core census protocol of the Forest Global Earth Observatory
{[}ForestGEO; REFS{]} collects the data required to calculate standing
dead wood, woody mortality, and delta.agb, but these have not been
calculated and reported for all sites for which the appropriate number
of censuses are available (n=1 for standing dead wood, n=2 for woody
mortality and delta.agb) {[}but see REFS{]}.

\textbf{A universal challenge in estimating biomass (living or dead)
from forest census data is applying appropriate allometries to convert
DBH measurements to biomass.} \emph{(Camille/Helene can write this
paragraph easily.)}

\subsection{Data reporting needs}

\textbf{We recommend that, unless they have some specific reason to do
otherwise, researchers calculate and report the values according to IPCC
standards:}

\begin{itemize}
\item
  adopt common standards for variables like min diameter of deadwood,
  select soil sampling increments to include a cutoff at 30.
\item
  report 95\% CIs, SE, or STD and n
\item
  report C variables in article text, table, or SI table. EFDB cannot
  accept data digitized from figures
\end{itemize}

\subsection{Database needs}

There are plenty of relevant, published data that are not included in
ForC. Systematic review of the literature could vastly improve data
coverage. \emph{(There are some efforts underway, including a few that
Susan can specify.)}

\subsection{IPCC}

An important challenge is that forests are changing rapidly, and data
collected a decaade ago may no longer be relevant, particularly in the
cases of C increments and fluxes.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{(fill this in)} %% optional section

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
Thank you to all researchers who collected and published the data
contained in ForC, and to all research assistants and collaborators who
have helped to build the database. Funding for this study was provided
by The Nature Conservancy, the Institute for Global Environmental
Strategies, WLS(?)
\end{acknowledgements}
